---
title: "Testing different normalisation schemes in the microbiome data set"
author: "Ricard Argelaguet"
date: "2/17/2020"
output: BiocStyle::html_document
---

```{r, message=FALSE}
library(tidyverse)
library(compositions)
library(reshape2)
# library(readr)
library(DESeq2)
library(cowplot)
library(MOFA2)
```

# Load data

```{r message=FALSE}
load("/Users/ricard/data/mofa_microbiome/rds/source_data_MOFA_microbiome.RDS")

metadata <- read.table("/Users/ricard/data/mofa_microbiome/metadata.txt", header=T)
```


Preprocess
```{r}
count_v <- viruses %>%
  group_by(sample, feature) %>%
  summarize(counts=sum(num_reads)) %>%
  ungroup %>%
  filter(complete.cases(.)) %>%
  dcast(sample ~ feature, value.var="counts", fill=0) %>%
  column_to_rownames("sample") %>%
  as.matrix

count_v[1:3,1:3]
```

Plot summary statistics
```{r}
hist(x = apply(count_v,1,sum),main="Sum of counts per sample")
hist(x = apply(count_v,2,sum),main="Sum of counts per virus")
```

```{r}
tail(sort(apply(count_v,2,sum)),n=5)
sort(apply(count_v,2,sum))
```


# Normalise virus data

## Normalise with CLR
```{r}
norm_v.clr <- as.data.frame(clr(count_v+0.1)) %>% t
norm_v.clr[1:3,1:3]
```


## Normalise with scran

```{r}
library(SingleCellExperiment)
library(scran)

sce <- SingleCellExperiment(list(counts=t(count_v)))
sce <- normalize(sce)

norm_v.scran <- logcounts(sce)
norm_v.scran[1:3,1:3]
```

## Normalise with DESeq2

Does not work because there are too many zeros
```{r}
# # Create a dummy coldata
# coldata <- DataFrame( foo=factor(rep(1,nrow(count_v))) )
# 
# # create DESeq object
# dds <- DESeqDataSetFromMatrix(t(count_v), colData=coldata, design=~1)
# 
# # VST
# dds <- varianceStabilizingTransformation(dds, blind=TRUE) 
# 
# # Fetch normalised counts
# norm_v.scran <- assay(dds)
# 
# norm_v.scran[1:3,1:3]
```

## Compare

```{r}
plot(
  x = colMeans(norm_v.scran), 
  y = colMeans(norm_v.clr), 
  xlab = "scran (mean expr per cell)", 
  ylab = "CLR (mean expr per cell)"
)
```

```{r}
plot(
  x = rowMeans(norm_v.scran), 
  y = rowMeans(norm_v.clr), 
  xlab = "scran (mean expr per gene)", 
  ylab = "CLR (mean expr per gene)"
)
```

```{r}
hist(
  x = diag(cor(norm_v.scran, norm_v.clr)), 
  xlim = c(0,1), 
  main="Correlation coefficient per gene (across cells)"
)
```

```{r}
hist(
  x = diag(cor(t(norm_v.scran), t(norm_v.clr))), 
  xlim=c(0,1), 
  main="Correlation coefficient per cell (across genes)"
)
```

# Run MOFA

## with CLR normalisation

```{r message=FALSE}
mofa.clr <- create_mofa(list("virus (CLR)"=norm_v.clr))

model_opts <- get_default_model_options(mofa.clr)
model_opts$num_factors <- 5

mofa.clr <- prepare_mofa(mofa.clr, model_options = model_opts)

mofa.clr <- run_mofa(mofa.clr)
```

## with scran normalisation

```{r message=FALSE}
mofa.scran <- create_mofa(list("virus (scran)"=norm_v.scran))

model_opts <- get_default_model_options(mofa.scran)
model_opts$num_factors <- 5

mofa.scran <- prepare_mofa(mofa.scran, model_options = model_opts)

mofa.scran <- run_mofa(mofa.scran)
```

## Add sample metadata to the models

```{r}
# makes sure that samples are in the correct order
metadata <- metadata[match(colnames(norm_v.scran), metadata$sample),]

# add dummy group column 
metadata <- metadata %>% mutate(group = "group1")

# add metadata to the model
samples_metadata(mofa.scran) <- metadata 
samples_metadata(mofa.clr) <- metadata 
```

# Compare MOFA models

## Variance explained

```{r}
p1 <- plot_variance_explained(mofa.clr, plot_total = T)[[2]] +
  coord_cartesian(ylim=c(0,0.25))
p2 <- plot_variance_explained(mofa.scran, plot_total = T)[[2]] +
  coord_cartesian(ylim=c(0,0.25))

plot_grid(plotlist=list(p1,p2))
```

## Factors correlated with library size

```{r}
to.plot <- cor(
  x = get_factors(mofa.clr)[[1]], 
  y = colSums(mofa.clr@data[[1]][[1]])
)[,1] %>% unname %>% abs

barplot(to.plot, ylim = c(0,1))
```

```{r}
to.plot <- cor(
  x = get_factors(mofa.scran)[[1]], 
  y = colSums(mofa.clr@data[[1]][[1]])
)[,1] %>% unname %>% abs

barplot(to.plot, ylim = c(0,1))
```

## Correlation of Factors within the model

CLR
```{r message=FALSE}
plot_factor_cor(mofa.clr)
```

scran
```{r message=FALSE}
plot_factor_cor(mofa.scran)
```

## Correlation of Factors between the models

```{r}
compare_factors(list("deseq"=mofa.scran, "clr"=mofa.clr))
```

```{r}
# p1 <- plot_factors(mofa.clr, 
#   factors = c(1,2), 
#   color_by = "Category", 
#   dot_size = 2.5
# ) + labs(title="CLR") + theme(legend.position = "none")
# 
# p2 <- plot_factors(mofa.scran, 
#   factors = c(1,2), 
#   color_by = "Category", 
#   dot_size = 2.5
# ) + labs(title="scran") + theme(legend.position = "none")
# 
# plot_grid(plotlist=list(p1,p2))
```

## Dimensionality reduction

Run UMAP
```{r}
set.seed(42)
mofa.clr <- run_umap(mofa.clr)
mofa.scran <- run_umap(mofa.scran)
```

Plot UMAP
```{r}
p1 <- plot_dimred(mofa.clr, method="UMAP", color_by = "Category", dot_size = 2.5) +
  labs(title="CLR") + theme(legend.position = "none") 
p2 <- plot_dimred(mofa.scran, method="UMAP", color_by = "Category", dot_size = 2.5) +
  labs(title="scran") + theme(legend.position = "none") 

plot_grid(plotlist=list(p1,p2))
```

# sessionInfo
```{r}
sessionInfo()
```

