---
title: "Testing different normalisation schemes in the microbiome data set"
author: "Ricard Argelaguet"
date: "2/17/2020"
output: BiocStyle::html_document
---

```{r, message=FALSE}
library(tidyverse)
library(compositions)
library(reshape2)
# library(readr)
library(DESeq2)
library(cowplot)
library(MOFA2)
```

# Load data

```{r message=FALSE}
load("/Users/ricard/data/mofa_microbiome/rds/source_data_MOFA_microbiome.RDS")

metadata <- read.table("/Users/ricard/data/mofa_microbiome/metadata.txt", header=T)
```


Preprocess
```{r}
count_b <- bacteria %>%
  mutate(Genus = if_else(is.na(Genus), Family, Genus))%>%
  group_by(sample, Genus)%>%
  summarize(counts = sum(numseqs))%>%
  filter(!is.na(Genus)) %>%
  dcast(sample ~ Genus, value.var="counts") %>% 
  column_to_rownames("sample") %>%
  as.matrix

count_b[1:3,1:3]
```

# Normalise bacterial data

## Normalise with CLR
```{r}
norm_b.clr <- as.data.frame(clr(count_b+0.1)) %>% t
norm_b.clr[1:3,1:3]
```


## Normalise with DESeq2
```{r}
# Create a dummy coldata
coldata <- DataFrame( foo=factor(rep(1,nrow(count_b))) )

# create DESeq object
dds <- DESeqDataSetFromMatrix(t(count_b), colData=coldata, design=~1)

# VST
dds <- varianceStabilizingTransformation(dds, blind=TRUE) 

# Fetch normalised counts
norm_b.deseq <- assay(dds)

norm_b.deseq[1:3,1:3]
```

## Compare

```{r}
plot(
  x = colMeans(norm_b.deseq), 
  y = colMeans(norm_b.clr), 
  xlab = "DESeq2 (mean expr per cell)", 
  ylab = "CLR (mean expr per cell)"
)
```

```{r}
plot(
  x = rowMeans(norm_b.deseq), 
  y = rowMeans(norm_b.clr), 
  xlab = "DESeq2 (mean expr per gene)", 
  ylab = "CLR (mean expr per gene)"
)
```

```{r}
hist(
  x = diag(cor(norm_b.deseq, norm_b.clr)), 
  xlim = c(0,1), 
  main="Correlation coefficient per gene (across cells)"
)
```

```{r}
hist(
  x = diag(cor(t(norm_b.deseq), t(norm_b.clr))), 
  xlim=c(0,1), 
  main="Correlation coefficient per cell (across genes)"
)
```

# Run MOFA

## with CLR normalisation

```{r message=FALSE}
mofa.clr <- create_mofa(list("bacteria (CLR)"=norm_b.clr))

model_opts <- get_default_model_options(mofa.clr)
model_opts$num_factors <- 10

mofa.clr <- prepare_mofa(mofa.clr, model_options = model_opts)

mofa.clr <- run_mofa(mofa.clr)
```

## with DESeq2 normalisation

```{r message=FALSE}
mofa.deseq <- create_mofa(list("bacteria (DESeq2)"=norm_b.deseq))

model_opts <- get_default_model_options(mofa.deseq)
model_opts$num_factors <- 10

mofa.deseq <- prepare_mofa(mofa.deseq, model_options = model_opts)

mofa.deseq <- run_mofa(mofa.deseq)
```

## Add sample metadata to the models

```{r}
# makes sure that samples are in the correct order
metadata <- metadata[match(colnames(norm_b.deseq), metadata$sample),]

# add dummy group column 
metadata <- metadata %>% mutate(group = "group1")

# add metadata to the model
samples_metadata(mofa.deseq) <- metadata 
samples_metadata(mofa.clr) <- metadata 
```

# Compare MOFA models

## Variance explained

```{r}
p1 <- plot_variance_explained(mofa.clr, plot_total = T)[[2]] +
  coord_cartesian(ylim=c(0,0.75))
p2 <- plot_variance_explained(mofa.deseq, plot_total = T)[[2]] +
  coord_cartesian(ylim=c(0,0.75))

plot_grid(plotlist=list(p1,p2))
```

## Factors correlated with library size

```{r}
to.plot <- cor(
  x = get_factors(mofa.clr)[[1]], 
  y = colSums(mofa.clr@data[[1]][[1]])
)[,1] %>% unname %>% abs

barplot(to.plot, ylim = c(0,1))
```

```{r}
to.plot <- cor(
  x = get_factors(mofa.deseq)[[1]], 
  y = colSums(mofa.clr@data[[1]][[1]])
)[,1] %>% unname %>% abs

barplot(to.plot, ylim = c(0,1))
```

## Correlation of Factors within the model

CLR
```{r message=FALSE}
plot_factor_cor(mofa.clr)
```

DESeq2
```{r message=FALSE}
plot_factor_cor(mofa.deseq)
```

## Correlation of Factors between the models

```{r}
compare_factors(list("deseq"=mofa.deseq, "clr"=mofa.clr))
```

```{r}
# p1 <- plot_factors(mofa.clr, 
#   factors = c(1,2), 
#   color_by = "Category", 
#   dot_size = 2.5
# ) + labs(title="CLR") + theme(legend.position = "none")
# 
# p2 <- plot_factors(mofa.deseq, 
#   factors = c(1,2), 
#   color_by = "Category", 
#   dot_size = 2.5
# ) + labs(title="DESeq2") + theme(legend.position = "none")
# 
# plot_grid(plotlist=list(p1,p2))
```

## Dimensionality reduction

Run UMAP
```{r}
set.seed(42)
mofa.clr <- run_umap(mofa.clr)
mofa.deseq <- run_umap(mofa.deseq)
```

Plot UMAP
```{r}
p1 <- plot_dimred(mofa.clr, method="UMAP", color_by = "Category", dot_size = 2.5) +
  labs(title="CLR") + theme(legend.position = "none") 
p2 <- plot_dimred(mofa.deseq, method="UMAP", color_by = "Category", dot_size = 2.5) +
  labs(title="DESeq2") + theme(legend.position = "none") 

plot_grid(plotlist=list(p1,p2))
```

# sessionInfo
```{r}
sessionInfo()
```

